---
title: "Permutation Test"
description:
  Small Simulation Study - Permutation Test of Sex Discrimination Data  
author: Ameya Teli
date: November 8, 2024
format: html
execute: 
  warning: false
  message: false
---
Source: openintro package
[R/data-sex_discrimination.R](https://github.com/OpenIntroStat/openintro/blob/HEAD/R/data-sex_discrimination.R)

Objective: 
See if there is a correlation between gender and being promoted from a study in the 1970s about bank manager recommendations based on sex. We want to show that the proportion of women who are promoted is less than men and that this is significant.  

Our null hypothesis is that gender does not affect the likelihood that men or women are promoted. 

```{r}

library(openintro)


sex_discrimination

```
The following code is to calculate the summary statistic for our data. 
```{r}
library(dplyr)
#the number of males and females in the study
nmale = sum(sex_discrimination$sex == "male")
nfemale = sum(sex_discrimination$sex =="female")

#the number of males and females that were promoted

males <- filter(sex_discrimination, sex == "male")

male_promoted = sum(males$decision == "promoted")

females <- filter(sex_discrimination, sex == "female")

female_promoted = sum(females$decision == "promoted")

#calculating the proportion of promotion with regard to gender 

prop_male <- male_promoted/nmale

prop_female <- female_promoted/nfemale

#create a dataframe with these summary statistics

gender_data <- data_frame(
  Gender = c("male", "female"), 
  Count = c(nmale, nfemale),
  Promoted = c(male_promoted, female_promoted), 
  Proportion = c(prop_male, prop_female)
)

gender_data

#Calculating the different between the proportion of males that were promoted and females that were promoted
obs_diff <- prop_male - prop_female
obs_diff
```
Now we generate a null sampling distribution 

```{r}
library(dplyr)
library(purrr)

#function to find the proportion of a specific gender which are given a promotion 

find_prop <- function(gender, data, promote_col){
  gender_specific <- filter(data, sex == gender)
  n_gender <- sum(data$sex == gender)
  gender_promoted <- sum(gender_specific[[promote_col]] == "promoted" )
  return(gender_promoted/n_gender)
}

#test that find_prop works 
find_prop("male", sex_discrimination, "decision")

#we care about whether or not females are promoted the same as or less than the males so we care about the sign of the difference (hence no absolute value)
find_prop_diff <- function(data, promote_col){
  find_prop("male", data, promote_col) - find_prop("female", data, promote_col)
}

find_prop_diff(sex_discrimination, "decision")

#function to permute the data & calculate the the proportion difference 
perm_data <- function(data){
  data <- data |>
    mutate(promote_perm = sample(decision, replace = FALSE)) 
  summarize(data, obs_prop_diff = find_prop_diff(data, "promote_perm"))
}

```

```{r}
library(ggplot2)

#mapping the permutation 
replication <- function(rep, data){
  map_dfr(1:rep, ~perm_data(data))
}

set.seed(4747)
perm_stats <- replication(100, sex_discrimination)

ggplot(perm_stats, aes(x = obs_prop_diff)) + 
  geom_histogram(binwidth = .05) + 
  geom_vline(aes(xintercept = obs_diff), color = "red") +
  labs(title = "Null Distribution of Promotion Differences by Gender", 
       x = "Proportion Difference (Permuted Data)",
       y = "Frequency") + 
  theme_minimal()

```

This is a graph of the null sampling distribution, that is we permuted the data many times switching the labels for men and women and promoted vs. not promoted and plotted this on the above histogram. The red vertical line shows our observed difference from the initial the sex_discrimination data from the study.  

```{r}
sum(perm_stats>=obs_diff)
```
There are only two entries from our null sampling distribution that are greater than or equal to our observed difference from the sampling distribution. 
This gives us a p value of 0.02 which is very small. Small enough for us to reject our null hypothesis. Thus the gender of the person does make a difference on whether or not they were promoted. More specifically to the point of this simulation study, the proportion of females promoted was significantly fewer than the proportion of men who were promoted. Thus gender, specifically being a female plays a significant role. 
