---
title: "Overview of DS002 Project 2"
description: "Shakespeare Textual Analysis: Romeo and Juliet"
author: "Ameya Teli"
subtitle: "December 9, 2024"
format:
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    embed-resources: true
    theme: serif
execute:
  echo: true
  warning: false
  message: false
---

## Project 2: Shakespeare Textual Analysis 

* This project is a textual analysis of Romeo and Juliet 
* It is split into a few stages, 3 of which are: 
  1. How much is death mentioned? 
  2. Punctuation
  3. Act III Analysis
  
#### Source 
::: {.class-name style="font-size: 50%;"}
This data was curated by [Nicola Rennie](https://github.com/nrennie) from The Complete Works of William Shakespeare (https://shakespeare.mit.edu), the Web's first edition of the Complete Works of William Shakespeare.
:::


## First, The Data
```{r, echo=FALSE}
romeo_juliet <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-09-17/romeo_juliet.csv')

head(romeo_juliet)
```

  
# How much is death mentioned? 

## Which character? 
```{r, echo=FALSE}
library(stringr)
library(dplyr)
library(tidyr)
library(ggplot2)

deathrj <- romeo_juliet |>
 filter(str_detect(dialogue, "death")) 
  

ggplot(deathrj, aes(x = character)) +
  geom_bar(stat = "count") + 
  labs(title = "Which character mentions death the most?", x = "Character") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```


## Mentions in Acts
```{r, echo=FALSE}
act_num <- deathrj |>
  mutate(act = str_extract(act, "(?<=\\s)\\S+")) 

ggplot(act_num, aes(x = act)) + 
  geom_bar(stat = "count") + 
  labs(title = "Which act has most mentions of death?", x = "Act") 

```

## Mentions in Acts 
* Most mentions of death in Act III and Act V
* A few notable deaths occur here 
* Maybe characters announce that they are going to die
* Other's talk about their deaths 

# Punctuation 

## Punctuation
* In the following plot I looked at places in the dialogue with something I called "strong punctuation" which was either "!" or "?".
* Using regular expressions 
  * str_detect(dialogue, "!+|\\?+")
* Created a new column (strong_punctuation) that was either true or false 


## Bar Chart
```{r, echo=FALSE}
filtered_act_num <- act_num |>
  filter(act == "III")|>
  mutate(strong_punctuation = str_detect(dialogue, "!+|\\?+")) |> #detecting any of those punctuation marks 
  mutate(scene = str_extract(scene, "(?<=\\s)\\S+"))  #extracting only the scene number, not the word "Scene" 
  

ggplot(filtered_act_num, aes(x = scene, fill = strong_punctuation))+
  geom_bar(position = "fill") + 
  labs(title = "Proportion of strong punctuation marks across scenes", x = "Scene", y = "Proportion")
```

## Bar Chart 
* Greater proportion of strong punctuation in Scene 5 than the rest
* Shows the prorportion of lines of dialogue that have strong punctuation versus lines of dialogue without 
* Stronger emotions related to the phrases that include a strong punctuation mark 
* Important events may occur in places with stronger emotions  

# Act III 

## Word/Character After Death in Act III

* Because Act III has the most mentions of death I wanted to look through the act to looking at the word/character after death    
* In this act two notable characters die: Tybalt and Mercutio
* Looking at the words/characters around the word "death" can tell us a lot about a moment 

* By creating a new df called filtered_act_num, I created a column called word_after_death using:
  * str_extract(dialogue, "death\\s*\\S+")
  
## Word/Character
```{r, echo=FALSE}
filtered_act_num <- act_num |>
  filter(act == "III")|>
  mutate(word_after_death = str_extract(dialogue, "death\\s*\\S+")) #extracting moments when there is a word/character after death using spaces

ggplot(filtered_act_num, aes(x = word_after_death))+
  geom_bar(stat = "count")+
  labs(title = "Death and the Word/Character after in Act III", 
       x = "Death + Word", 
       y = "Count")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Thank you! 
