---
title: "Overview of DS002 Projects"
author: "Ameya Teli"
subtitle: "December 9, 2024"
format:
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Project 1: Data Visualizations 
For this project one of the data sets I worked with was about the London Marathon from 1980 to 2020.
 
* London Marathon: contains data about applicants, accepted runners, starters, finishers, and data on winners in the elite section 


```{r, echo=FALSE}
london_marathon <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-04-25/london_marathon.csv')
soccer <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-04-04/soccer21-22.csv')

```

```{r}
head(london_marathon)
```

## London Marathon 
 
```{r, echo = FALSE}
library(ggplot2)
library(tidyverse)


ggplot(data = london_marathon, 
       aes(
         x=Year, 
         y=Accepted)
       ) +
  geom_point() + 
  labs(title ="Number of Accepted Runners to the London Marathon vs Year")

```
[This is the shiny application to plot data in comparison to the number of runners accepted to the London Marathon](http://ameyateli.shinyapps.io/londonmarathon)

## Project 2: Shakespeare Textual Analysis 

* This project is a textual analysis of Romeo and Juliet 
* The word of interest for most of this analysis is "death" given that Romeo and Juliet is a tragedy 
* 

```{r, echo=FALSE}

```



